Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-08-05 agujournaltemplate.tex
Post-Ago comments, just a few small things remain before final checks

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at depths of 1, 7, 15, 25, 35, 55, 75, and 100 m in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts that are shown in Figure \ref{fig:zscore}), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}). A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera. The identified genera were further grouped into functional groups based on their taxonomic classification in WoRMS.

@@

\subsection{Community reorganization and biomass shifts}
    % P1: Introductory paragraph highlighting the overall dynamics/shifts
    This study shows  a multi-annual oscillatory nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
    These changes are clearly visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustcomp}). After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all in coincidence with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014. 
    By considering the phytoplankton community time series up to the end of 2017, the temporally complete dataset, we found that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}. In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations (Figure \ref{fig:zscore}). Concurrently, we observed an increase in chlorophyll \textit{a} (Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (Figure \ref{fig:clustering} $a$ and $c$).

    % P2: Discuss community reorganisation, biomass shifts vs HPLC data 
    When comparing the two clusters, we observed a marked reduction in cell counts, which corresponded to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}). The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community might have transitioned to smaller cells, which were probably below the detection size limit of optical microscopes \cite{muller-karger_scientific_2019}. This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}. Increases in temperature can lead to a decrease in the relative contribution of larger cells in the community, irrespective of nutrient availability, but shifts towards smaller cells are particularly prevalent when temperatures increase and nutrient supply decreases \cite{mousing_global_2014}, as we observed in the Cariaco Basin.

    % Disucssion of HPLC vs Microscopy!
    A similar general pattern emerged when comparing our results with those based on High-Performance Liquid Chromatography (HPLC) data from the Cariaco Basin, which were analyzed and presented by \citeA{pinckney_phytoplankton_2015}. However, some differences emerged between our results and those obtained with the HPLC data. The HPLC data showed an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}. It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006. Total chlorophyll \textit{a} could not be quantified with a consistent method across the different laboratories \cite{pinckney_phytoplankton_2015}. This is the reason why we focused our analysis on the more consistent and temporally complete microscopy cell count and fluorometric chlorophyll \textit{a} data.
    Consistent with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, showed a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
    Additionally, despite cell counts of all functional groups decreased, the HPLC data indicated an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}. This provides a further strong indication that the phytoplankton community shifted to smaller sizes, an aspect that was not detected by previous studies using light microscopy data.

@@

    % P5: Climate impacts
    Although regional effects may vary, a negative phase in the AMO is generally associated with periods of colder temperatures and a southward shift of the ITCZ \cite{knight_climate_2006, colna2017latitudinal}. In the Cariaco Basin, a southward shift of the ITCZ correlates with increased wind-driven upwelling \cite{taylor_ecosystem_2012}. The threshold responses between functional groups, which relate to shifts in cell count abundances, showed contrasting patterns. Diatoms, dinoflagellates, and haptophytes exhibited strong response thresholds for negative AMO anomalies, which pointed to the influence of stronger upwelling conditions. The abundance of haptophytes, additionally, showed a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} based on HPLC data showing haptophytes blooming predominantly above the mixed layer and at a lower frequency than other functional groups.

@@

    % 1.  Discuss Limitations of Light microscopy cell counts in CARIACO Time Series
    Using microscopy cell counts as a basis for our analyses allowed us to focus on changes in the microphytoplankton community. Our choice of grouping cell counts by genus was based on the assumption that species within a genus generally exhibit similar eco-physiological responses. The output of the Gradient Forest model was further grouped by broad functional types to facilitate the identification of shifts and trends. The phytoplankton genera that could be best predicted were mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$). This does not represent a complete picture of the phytoplankton community of the CARIACO basin. The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} captured all size classes of phytoplankton but, as we mentioned above, it had a limited temporal coverage and the dataset was assembled by four different laboratories. With the caveat that the microscopy counts captured microphytoplankton almost exclusively, the discrepancies with the HPLC pigment analysis can be explained. Drastic declines in cell counts reflected shifts to smaller species, an aspect consistent with size spectra calculated by previous studies using HPLC pigment data \cite{lorenzoni_characterization_2015}.

    % 2. Discuss limitations of GF analysis & Limitation to bottom-up drivers and climate indices
    The limited scope of the microscopy cell counts also affected the conclusions that could be drawn from the Gradient Forest analysis, as these results are limited to microphytoplankton. 
    Another potential caveat of our methodology is that many of the predictors in ecological time series in general and, specifically, in this study, are correlated (Figure \ref{fig:sup:correlation} in Appendix). This can lead to the identification of a correlated predictor that has no direct effect on community changes. However, the Gradient Forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can partly mitigate the problem \cite{ellis_gradient_2012}. Additionally, random forest models do not consider temporal relationships between data points, which is why we included tests for the predictive relationships with time lags of our predictors. Without adjusting for time lags, SST and nitrate concentration are the strongest predictors of community change (Figure \ref{fig:sup:GFoutput_nolags} in Appendix). Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both the SST and the AMO index yielded a relatively low value around \qty{0.03}{}. This is expected given the noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the Gradient Forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}. It is important to consider that although predictive relationships between drivers and ecological variables are identified, this does not necessarily imply a mechanistic causal relationship . The scope of our analysis, like any statistical analysis, is exploratory and aims at narrowing down potential key relationships. Mathematical modeling and experimental work are required to develop a clearer mechanistic understanding of the drivers of ecosystem change in the Cariaco Basin.

@@

\section{Conclusions}
    We found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin. Specifically, microscopy cell counts revealed the presence of two distinct clusters defined by high- and low-upwelling conditions that corresponded to anomalies in the AMO index. The community changes did not indicate a permanent regime shift, a return to the early cluster was observed towards the end of the time series. We observed no significant changes in diversity metrics between the two clusters, except for lower turnover and higher community evenness under low-upwelling conditions.
    Despite multi-annual trends in warming, reduced productivity, and drastic declines in micro-phytoplankton cell counts, the phytoplankton community returned to previous levels once the long-term climatic and local physical conditions were restored. 
    The strongest variable predicting changes in the phytoplankton community was the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles. We concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state for the phytoplankton community of the Cariaco Basin. 
    Notwithstanding a number of limitations, our analysis points to the incipient return of the Cariaco Basin to a pre-2004 regime, hinting at an oscillatory nature of phytoplankton communities under global change, an aspect that emphasizes the relevance of long-term time series and deserves further attention. 
    Future studies, ideally based on mathematical modeling, could focus on the relative impact of top-down vs. bottom-up drivers,  aspects remaining largely unexplored in the Cariaco Basin.

@@

%!
\section*{Open Research Section}
All data used in this manuscript are publicly available through open access. The CARIACO time series data we used was accessed from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) under the doi \url{10.1575/1912/bco-dmo.3095.1}. This includes the Niskin bottle dataset for monthly in-situ measurements of chlorophyll \textit{a} and nutrient data \url{https://www.bco-dmo.org/dataset/3093}, the CTD dataset \url{https://www.bco-dmo.org/dataset/3092}, and the phytoplankton taxonomy dataset \url{https://www.bco-dmo.org/dataset/3095}.
The climate variables, such as wind speed, precipitation and evaporation, were extracted from the "ERA5 monthly averaged data on single levels from 1940 to present" dataset \url{https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7}.
The AMO index was retrieved from the Climate Data Guide provided by the NSF National Center for Atmospheric Research at \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} and the MEI v.2 was accessed from NOAA Physical Sciences Laboratory (PSL) at \url{https://psl.noaa.gov/enso/mei/}.
All the code and data used in this analysis and to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
%TODO add Zenodo link before submission

>>>>

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at depths of 1, 7, 15, 25, 35, 55, 75, and 100 m in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. 
    %from Mutshinda et al. 2013
    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts that are shown in Figure \ref{fig:zscore}), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}). A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera. The identified genera were further grouped into functional groups based on their taxonomic classification in WoRMS.

@@

\subsection{Community reorganization and biomass shifts}
    % P1: Introductory paragraph highlighting the overall dynamics/shifts
    This study shows the multi-annual oscillatory nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
    These changes are clearly visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustcomp}). After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014. 
    By considering the phytoplankton community time series up to the end of 2017, the temporally complete dataset, we found that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}. In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations (Figure \ref{fig:zscore}). Concurrently, we observed an increase in chlorophyll \textit{a} (Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (Figure \ref{fig:clustering} $a$ and $c$).

    % P2: Discuss community reorganisation, biomass shifts vs HPLC data 
    When comparing the two clusters, we observed a marked reduction in cell counts, which corresponded to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}). The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community might have transitioned to smaller cells, which were probably below the detection size limit of optical microscopes \cite{muller-karger_scientific_2019}. This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}. Increases in temperature can lead to a decrease in the relative contribution of larger cells in the community, irrespective of nutrient availability, but shifts towards smaller cells are particularly prevalent when temperatures increase and nutrient supply decreases \cite{acevedo-trejos_glimpse_2014, mousing_global_2014}, as we observed in the Cariaco Basin.

    % Disucssion of HPLC vs Microscopy!
    A similar general pattern emerged when comparing our results with those based on High-Performance Liquid Chromatography (HPLC) data from the Cariaco Basin, which were analyzed and presented by \citeA{pinckney_phytoplankton_2015}. However, some differences emerged between our results and those obtained with the HPLC data. The HPLC data showed an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}. It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006. Total chlorophyll \textit{a} could not be quantified with a consistent method across the different laboratories \cite{pinckney_phytoplankton_2015}. This is the reason why we focused our analysis on the more consistent and temporally complete microscopy cell count and fluorometric chlorophyll \textit{a} data.
    Consistent with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, showed a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
    Additionally, despite decreased cell counts of all functional groupsing, the HPLC data indicated an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}. This provides a further strong indication that the phytoplankton community shifted to smaller cell sizes, which were not detected using light microscopy.

@@

    % P5: Climate impacts
    Although regional effects may vary, a negative phase in the AMO is generally associated with periods of colder temperatures and a southward shift of the ITCZ \cite{knight_climate_2006, colna2017latitudinal}. In the Cariaco Basin, a southward shift of the ITCZ correlates with increased wind-driven upwelling \cite{taylor_ecosystem_2012}. The threshold responses between functional groups, which relate to shifts in cell count abundances, showed contrasting patterns. Diatoms, dinoflagellates, and haptophytes exhibited strong response thresholds for negative AMO anomalies, which pointed to the influence of stronger upwelling conditions. The abundance of haptophytes, additionally, showed a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} based on HPLC data showing haptophytes blooming more frequently during the time period of 2006 to 2010, which was characterized by mostly positive AMO anomalies.

@@

    % 1.  Discuss Limitations of Light microscopy cell counts in CARIACO Time Series
    Using microscopy cell counts as a basis for our analyses allowed us to focus on changes in the microphytoplankton community. Our choice of grouping cell counts by genus was based on the assumption that species within a genus generally exhibit similar eco-physiological responses. The output of the Gradient Forest model was further grouped by broad functional types to facilitate the identification of shifts and trends. The phytoplankton genera that could be best predicted were mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$). This does not represent a complete picture of the phytoplankton community of the Cariaco basin. The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} captured all size classes of phytoplankton but, as we mentioned above, it had a limited temporal coverage and the dataset was assembled by four different laboratories. With the caveat that the microscopy counts captured microphytoplankton almost exclusively, the discrepancies with the HPLC pigment analysis can be explained. Drastic declines in cell counts reflected shifts to smaller species, an aspect consistent with size spectra calculated by previous studies using HPLC pigment data \cite{lorenzoni_characterization_2015}.

    % 2. Discuss limitations of GF analysis & Limitation to bottom-up drivers and climate indices
    The limited scope of the microscopy cell counts also affected the conclusions that could be drawn from the Gradient Forest analysis, as these results are limited to microphytoplankton. 
    Another potential caveat of our methodology is that many of the predictors in ecological time series in general and, specifically, in this study, are correlated (Figure \ref{fig:sup:correlation} in Appendix). This can lead to the identification of a correlated predictor that has no direct effect on community changes. However, the Gradient Forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can partly mitigate this problem \cite{ellis_gradient_2012}. Additionally, random forest models do not consider temporal relationships between data points, which is why we included tests for the predictive relationships with time lags of our predictors. Without adjusting for time lags, SST and nitrate concentration are the strongest predictors of community change (Figure \ref{fig:sup:GFoutput_nolags} in Appendix). Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and the AMO index yielded a relatively low value around \qty{0.03}{}. This is expected given the noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the Gradient Forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}. It is important to consider that although predictive relationships between drivers and ecological variables are identified, this does not necessarily imply a mechanistic causal relationship. The scope of our analysis, like any statistical analysis, is exploratory and aims at narrowing down potential key relationships. Mathematical modeling and experimental work are required to develop a clearer mechanistic understanding of the drivers of ecosystem change in the Cariaco Basin.

@@

\section{Conclusions}
    We found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin. Specifically, microscopy cell counts revealed the presence of two distinct clusters defined by high- and low-upwelling conditions that corresponded to anomalies in the AMO index. The community changes did not indicate a permanent regime shift, since a return to the early cluster was observed towards the end of the time series. We observed no significant changes in diversity metrics between the two clusters, except for lower turnover and higher community evenness under low-upwelling conditions.
    Despite multi-annual trends in warming, reduced productivity, and drastic declines in micro-phytoplankton cell counts, the phytoplankton community returned to previous levels once the long-term climatic and local physical conditions were restored. 
    The strongest variable predicting changes in the phytoplankton community was the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles. We concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state for the phytoplankton community of the Cariaco Basin. 
    Notwithstanding a number of limitations, our analysis points to the incipient return of the Cariaco Basin to a pre-2004 regime, hinting at an oscillatory nature of phytoplankton communities under global change, an aspect that emphasizes the relevance of long-term time series and deserves further attention. 
    Future studies, ideally based on mathematical modeling, could focus on the relative impact of top-down vs. bottom-up drivers,  aspects remaining largely unexplored in the Cariaco Basin.

@@

%!
\section*{Open Research Section}
All data used in this manuscript are publicly available. The CARIACO time series data we used was accessed from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) under the doi \url{10.1575/1912/bco-dmo.3095.1}. This includes the Niskin bottle dataset for monthly in-situ measurements of chlorophyll \textit{a} and nutrient data \url{https://www.bco-dmo.org/dataset/3093}, the CTD dataset \url{https://www.bco-dmo.org/dataset/3092}, and the phytoplankton taxonomy dataset \url{https://www.bco-dmo.org/dataset/3095}.
The climate variables, such as wind speed, precipitation and evaporation, were extracted from the "ERA5 monthly averaged data on single levels from 1940 to present" dataset \url{https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7}.
The AMO index was retrieved from the Climate Data Guide provided by the NSF National Center for Atmospheric Research at \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} and the MEI v.2 was accessed from NOAA Physical Sciences Laboratory (PSL) at \url{https://psl.noaa.gov/enso/mei/}.
All the code and data used in this analysis and to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
%TODO add Zenodo link before submission
